Assessing Power of the California Microhap Baseline for GSI (and calculating Fst)

### 2024-05-31 002-GSI_and_Fst.Rmd
Out locus summaries by collection

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    final_baseline = "data/subRoSA_baseline_rubias_FEB2024_w_mixture_revised.csv",
    final_baseline_apr = "data/subRoSA_baseline_rubias_APR2024_w_mixture_revised_rosa_added.csv",
    lfar_genos = "data/metagenos_hap2col_lfar_all_samps.csv",
    pop_labels = "inputs/reference-collection-names.csv",
    map_notations = "inputs/map-notations.tsv"
  )
  # outputs:
  output_list <- list(
    assig_table = "results/GSI_and_Fst/assignment-table-full-baseline.pdf",
    fst_table = "results/GSI_and_Fst/full-baseline-but-no-lfar-Fst-table.pdf",
    assig_table_lfar_all = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar.pdf",
    ass_table80 = "results/GSI_and_Fst/ass-table-80.pdf",
    rosa_gsi_table = "results/GSI_and_Fst/rosa-gsi-table.pdf",
    tex_table = "tex/inputs/samples-table.tex",
    gsi_fst_fig = "results/GSI_and_Fst/gsi_and_fst_fig.pdf",
    gsi_fst_fig_crop = "results/GSI_and_Fst/gsi_and_fst_fig-crop.pdf",
    gsi_fst_fig_tex = "tex/images/gsi_and_fst_fig-crop.pdf",
    ass_table80_crop = "results/GSI_and_Fst/ass-table-80-crop.pdf",
    ass_table80_tex = "tex/images/ass-table-80-crop.pdf",
    rosa_gsi_table_crop = "results/GSI_and_Fst/rosa-gsi-table-crop.pdf",
    rosa_gsi_table_tex = "tex/images/rosa-gsi-table-crop.pdf"
  )
}

@@

@@

# we will now use that min_diploid number to downsample
# each collection randomly, without replacement 1000 times and 
# calculate the mean number of alleles.
sub_sampled_locus_num_alle <- fbl_no_miss %>%
  left_join(min_gene_copies, by = join_by(locus)) %>%
  group_by(repunit, collection, locus) %>%
  summarise(
    mean_num_alleles = list(subsample_num_alleles(value, min_gc[1], N = 100))
  ) %>%
  ungroup() %>%
  unnest(mean_num_alleles)

@@

>>>>

```{r}
if(exists("snakemake")) {
  input_list <- snakemake@input
  output_list <- snakemake@output
} else {
  input_list <- list(
    final_baseline = "data/subRoSA_baseline_rubias_FEB2024_w_mixture_revised.csv",
    final_baseline_apr = "data/subRoSA_baseline_rubias_APR2024_w_mixture_revised_rosa_added.csv",
    lfar_genos = "data/metagenos_hap2col_lfar_all_samps.csv",
    pop_labels = "inputs/reference-collection-names.csv",
    map_notations = "inputs/map-notations.tsv"
  )
  # outputs:
  output_list <- list(
    assig_table = "results/GSI_and_Fst/assignment-table-full-baseline.pdf",
    fst_table = "results/GSI_and_Fst/full-baseline-but-no-lfar-Fst-table.pdf",
    assig_table_lfar_all = "results/GSI_and_Fst/assignment-table-full-baseline-plus-lfar.pdf",
    ass_table80 = "results/GSI_and_Fst/ass-table-80.pdf",
    rosa_gsi_table = "results/GSI_and_Fst/rosa-gsi-table.pdf",
    tex_table = "tex/inputs/samples-table.tex",
    gsi_fst_fig = "results/GSI_and_Fst/gsi_and_fst_fig.pdf",
    gsi_fst_fig_crop = "results/GSI_and_Fst/gsi_and_fst_fig-crop.pdf",
    gsi_fst_fig_tex = "tex/images/gsi_and_fst_fig-crop.pdf",
    ass_table80_crop = "results/GSI_and_Fst/ass-table-80-crop.pdf",
    ass_table80_tex = "tex/images/ass-table-80-crop.pdf",
    rosa_gsi_table_crop = "results/GSI_and_Fst/rosa-gsi-table-crop.pdf",
    rosa_gsi_table_tex = "tex/images/rosa-gsi-table-crop.pdf",
    pop_gen_by_loc_coll = "tex/supp_data/Supp-Data-2-pop-gen-summaries-by-locus-and-collection.csv"
  )
}

@@

While we are at it, let's look at the distribution of the E2 alleles across
collections.
```{r}
# count up genos at each variant
marks <- paste("s", 1:12, sep = "")
exploded <- rosa_hapstrs %>%
  filter(!(collection %in% c("ColeRHsp", "SmithFall"))) %>%
  separate(hapstr, into = paste("s", 0:12, sep = ""), sep = "") %>%
  select(-s0)
rh_counts <-  exploded %>%
  pivot_longer(cols = starts_with("s"), names_to = "variant", values_to = "geno") %>%
  mutate(variant_f = factor(variant, levels = marks)) %>%
  count(repunit, collection, variant_f, geno)

# look just at s2, s3, and s6
rh_counts %>%
  filter(variant_f %in% c("s2", "s3", "s6")) %>%
  filter(geno %in% c("H", "W"))
```

So, there is only one more E2-het outside of the Sacto, and that is from TRHsp.

Let's go back and find that:
```{r}
exploded %>% 
  filter(repunit == "klamathspring" & s6 == "H") %>%
  left_join(rosa_hapstrs, by = join_by(indiv))

@@

# we will now use that min_diploid number to downsample
# each collection randomly, without replacement 1000 times and 
# calculate the mean number of alleles.
set.seed(1)
sub_sampled_locus_num_alle <- fbl_no_miss %>%
  left_join(min_gene_copies, by = join_by(locus)) %>%
  group_by(repunit, collection, locus) %>%
  summarise(
    mean_num_alleles = list(subsample_num_alleles(value, min_gc[1], N = 100))
  ) %>%
  ungroup() %>%
  unnest(mean_num_alleles)

@@

## Locus specific pop-gen summaries

Now, we also are going to make summaries for each locus in each collection.
Carlos wanted a supplemental data table that has Hexp and Hobs for each
collection.  I figure we might as well add the number of diploids and the
number of alleles in there as well.   All of that information is in
`locus_summaries` in a tidy format, but we will make it wide for this.
Carlos wanted to add it to the Supp-Data-1-Amplicon-info.xlsx, but I think
that is actually a little crazy because we are going to have so many columns in this
output. So, I will put it into a CSV file.
```{r}
by_locus_table <- locus_summaries %>%
  select(locus, collection, num_diploids, num_alleles, exp_Hz, obs_Hz) %>%
  pivot_wider(
    names_from = collection,
    values_from = num_diploids:obs_Hz,
    names_sep = "."
  )

write_csv(by_locus_table, file = output_list$pop_gen_by_loc_coll)
```
